---
title: "Systematic Literature Reviews with R:Session 2"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
date: "2026-04-08"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Systematic **Literature Review**

This workshop introduces how R can be used to support systematic review
workflows. The previous session provided an overview of the systematic
literature review process, including how to formulate research
questions, develop search strategies, and use R to import bibliographic
data from literature databases using revtools and synthesisr. It also
covered reference deduplication to prepare datasets for literature
screening and meta-analysis.

In today’s workshop, we will focus on the screening process. We will
demonstrate how to use revtools to screen titles, abstracts, and full
texts. In addition, we will introduce text mining, meta-analysis, and
visualising results in R.

## Set up

### Package set up

This part may vary if you are working locally but to get set up in
Noteable below you can find all the packages that need to be installed
and loaded to run the code.

```{r }
# Install and load the required packages
install.packages("revtools", dependencies = TRUE)
install.packages("synthesisr")
install.packages("cld3")
install.packages("widyr")
install.packages("metafor") 
library(revtools)
library(synthesisr)
library(stringr)
library(dplyr)
```

## Include and Exclude criteria prescreening:

Last week, we performed deduplication using both DOI and title + author
matching. It is also beneficial to use revtools(screen_duplicates) for a
final manual check, as automated methods may miss subtle duplicates
caused by variations in metadata (e.g., formatting differences, missing
DOIs, or slight changes in titles). A manual review step helps ensure
higher accuracy and reduces the risk of retaining duplicate records or
accidentally removing relevant studies.

```{r }
refdata <- read.csv("../Data/Clean/metadata.csv")
screen_duplicates(refdata)
```

### Language: English Only

Other laguage tool such textcat is not accurate as cld3, to improve the
accuracy we include both title and abstract column

```{r }
#you can import the deduplication data now;
refdata <- read.csv("../Data/Clean/cleaned.csv")

library(cld3)
refdata$language <- detect_language(paste(
  refdata$title,
  refdata$abstract
))
# keep a record for the literature is not recognised in english and review it before you remove it
noendata <- refdata %>%
  filter(language != "en")
english_data <- refdata %>%
  filter(language == "en")
```

### Year: 2016-2026

```{r }
# You can check the column type, if it is character type you can use as.numeric to convert it
str(english_data$year)
filtered_data <- english_data %>%
  filter(year >= 2016 & year <= 2026)
year_out <- english_data %>%
  filter(is.na(year) | year < 2016 | year > 2026)
```

### Keywords search prescreening search

Follow the same search strategy(("digital device*" OR smartphone* OR
"mobile phone*" OR "screen time" OR "social media" OR "internet use" OR
"digital technolog*" OR "online activity") AND ("mental health" OR
wellbeing OR "well-being" OR depression OR anxiety OR stress OR
"psychological distress" OR "mental disorder*") AND ("young adult*" OR
youth OR "college student*" OR "university student*" OR adolescent\*))

This step can help to reduce the task amount for large dataset as we
have to a smaller size, which is more handy for screening.

```{r }
# we can check if any empty title/abstract in our dataset
no_abstract <- filtered_data %>%
  filter(is.na(abstract) | abstract == "")
# we define the keywords follow the strategy we set from PICO/PEO, P:population, I:intervention/E:exposure,C:comparison, O:outcome. You can expand the key terms rather to strict by adding anonymous.

digital_terms <- c(
  "digital device", "smartphone", "mobile phone",
  "screen time", "social media", "internet",
  "online", "digital technolog", "device use"
)

mental_health_terms <- c(
  "mental health", "wellbeing", "well-being",
  "depression", "anxiety", "stress",
  "psychological", "distress", "mental disorder",
  "mood", "emotional"
)

population_terms <- c(
  "young adult", "youth", "student",
  "college", "university", "adolescent",
  "teen", "young people"
)
# Follow the  Boolean we use | for or in each keyword group
pattern_digital <- paste0(digital_terms, collapse = "|")
pattern_mental  <- paste0(mental_health_terms, collapse = "|")
pattern_pop     <- paste0(population_terms, collapse = "|")

# Adding exclude terms, depends on your familiarity with your data and make sure it not include any bias on viewing RQ.For this example data, We would like to remove the literature with animal studies
exclude_terms <- c("animal", "rat", "mouse","monkey")
pattern_exclude <- paste(exclude_terms, collapse = "|")

# Before we apply the keyword searching on our data, we will build a new column called text to store the lowercased title and abstract
with_abstract <- filtered_data %>%
  filter(!is.na(abstract) & abstract != "") %>%
  mutate(text = tolower(paste(title, abstract))) %>%
  filter(
    grepl(pattern_digital, text) &
    grepl(pattern_mental, text) &
    grepl(pattern_pop, text) &
    !grepl(pattern_exclude, text)
  )
combined_data <- bind_rows(with_abstract, no_abstract)
# Keep a record for the data has been filtered out
removed_data <- filtered_data %>%
  filter(!is.na(abstract) & abstract != "") %>%
  mutate(text = tolower(paste(title, abstract))) %>%
  filter(
    !(grepl(pattern_digital, text) &
    grepl(pattern_mental, text) &
    grepl(pattern_pop, text))|
    grepl(pattern_exclude, text)
  )
nrow(combined_data)
```

## Screening

### Interactive screen interface with r revtools screen title first

```{r }
screen_titles(combined_data)
```

### Abstract screen

```{r }
screen_abstracts(combined_data)
```

### Practice with random select data from big dataset

```{r }
#Random seek 50 from dataset
set.seed(123)
sample_50 <- combined_data %>%
  slice_sample(n = 50)
screen_titles(sample_50)
screen_abstracts(sample_50)
```

### manual review without revtools

```{r }
screentitle <- combined_data %>%
  select(title_clean)
screenabstract <- combined_data %>%
  select(abstract)
```

### Build a table/record for screening

```{r }
screening_record <- sample_50 %>%
  mutate(
    screening_decision = NA,  # include / exclude / unsure for options
    exclusion_reason = NA     # reason if excluded/note
  ) %>%
  select(
    title,
    author,
    year,
    doi,
    abstract,
    screening_decision,
    exclusion_reason
  )
```

```{r }
# manual type in for the reference
screening_record$screening_decision[1] <- "include"
screening_record$screening_decision[2] <- "exclude"
screening_record$exclusion_reason[2] <- "Not relevant population"
# Practice to write more records
```

### Combine the two reviewers’ assessments into a single dataset

Combine the two reviewers’ assessments into a single dataset, or use
library(readxl) to import the Excel data. This approach helps merge and
reconcile the screening results from both reviewers.

```{r }
# Two sample data from screening title
rev1 <- read.csv("../Data/Clean/sampletitle.csv")
rev2 <- read.csv("../Data/Clean/sampletitle2.csv")
# merge with title name
merged <- rev1 %>%
  full_join(rev2, by = "title_clean", suffix = c("_r1", "_r2"))
table(merged$screened_titles_r1, merged$screened_titles_r2)
# create a column to recorde agreement
merged <- merged %>%
  mutate(
    agreement = ifelse(screened_titles_r1 == screened_titles_r2,
                        "agree", "disagree")
  )
```

## META ANALYSIS

After screening full texts, you can extract your meta data in a table,
or manual build one in R. metafor package (gold standard in R
meta-analysis) <https://cran.r-project.org/package=metafor>
Meta-analysis visualisation (forest funnel plots)
<https://www.maths.bris.ac.uk/R/web/packages/metaviz/vignettes/metaviz.html>

```{r }
# load package
  library(metafor)

#Create an output folder 
if (!dir.exists("../output")) {
  dir.create("../output")
}

# Example dataset (manual import here as example, replace with your real dataset)
dat <- data.frame(
  study = c("Study1", "Study2", "Study3", "Study4", "Study5",
            "Study6", "Study7", "Study8"),
  
  yi = c(0.20, 0.35, -0.10, 0.50, 0.15, 0.40, 0.05, 0.25),  # effect sizes
  
  sei = c(0.10, 0.15, 0.20, 0.12, 0.18, 0.14, 0.16, 0.11)   # standard errors
)

```

Compute variance

```{r}
# Compute variance converts standard error into variance as Meta-analysis needs variance, not SE
dat$vi <- dat$sei^2
# Save cleaned dataset
write.csv(dat, "../output/meta_input_data.csv", row.names = FALSE)
```

Random-effects meta-analysis (REML)

```{r}
# Random-effects meta-analysis (REML)
res <- rma(yi = yi,
           vi = vi,
           data = dat,
           method = "REML")
# Print results
print(res)
# Save model summary (text output)
capture.output(summary(res), file = "../output/meta_analysis_summary.txt")
```

Extract results

```{r}
# Extract results
results_df <- data.frame(
  pooled_effect = res$b,
  ci_lower = res$ci.lb,
  ci_upper = res$ci.ub,
  I2 = res$I2,
  tau2 = res$tau2,
  Q_pvalue = res$QEp
)

write.csv(results_df, "../output/meta_results_summary.csv", row.names = FALSE)

```

Forest plot

```{r}
# Forest plot
png("../output/forest_plot.png", width = 1000, height = 700)
forest(
  res,
  slab = dat$study,
  xlab = "Effect Size",
  mlab = "Random Effects Model (REML)"
)
dev.off()

```

Funnel Plot

```{r}
# Funnel Plot 
png("../output/funnel_plot.png", width = 800, height = 700)
funnel(
  res,
  xlab = "Effect Size",
  ylab = "Standard Error"
)
dev.off()


```

Publication bias check Egger test

```{r}
# publication bias check egger test
egger <- regtest(res, model = "rma")
capture.output(egger, file = "../output/egger_test.txt")

```

Leave-one-out analysis

```{r}
# Leave-one-out analysis
loo <- leave1out(res)
write.csv(as.data.frame(loo), "../output/leave_one_out.csv", row.names = FALSE)


```

Influence diagnose

```{r}
# influence diagnose
png("../output/influence_plot.png", width = 900, height = 700)
inf <- influence(res)
plot(inf)
dev.off()
```

Interpretation random-effects result Summary: A random-effects
meta-analysis revealed a significant positive effect (β = 0.25, p \<
.001) with low-to-moderate heterogeneity (I² = 33.5%), indicating
reasonably consistent effects across studies. egger test: A regression
test for funnel plot asymmetry did not indicate significant small-study
effects (z = -1.37, p = .171), suggesting no strong evidence of
publication bias. The estimated intercept implied a positive effect
among highly precise studies (b = 0.63, 95% CI [0.08, 1.18]). A
leave-one-out sensitivity analysis indicated that the overall effect
size remained stable (β ranged from 0.22 to 0.28) and statistically
significant across all iterations (p \< .001). Heterogeneity estimates
showed minor variation but no evidence that any single study
disproportionately influenced the results, supporting the robustness of
the meta-analytic findings.

## Ploting for the reference

factoextra (for clustering + PCA plots)
<https://rpkgs.datanovia.com/factoextra/> ggplot2 official learning
<https://ggplot2.tidyverse.org/>

### Word clustering is applied to titles, as abstracts are longer and

require more processing; titles and keywords are preferred.

Cluster Analysis in R <https://rpkgs.datanovia.com/factoextra/>

Cluster analysis tutorials:
<https://www.r-bloggers.com/2021/10/cluster-analysis-using-r-2/>

Install the missing library and update the existing libraries to the
right version

Before running the below you will have to go to Session\> Restart R.

This is because the rlang version that come with Noteable it is not
compatible with factor extra so we first need to get the right version
of it.

You will also probably see a prompt that will ask you to reboot R, press
yes

```{r}
install.packages("rlang", dependencies = TRUE) # this is needed because the version in Noteable is too old
install.packages("factoextra")
```

Make sure that all the needed libraries are loaded

```{r }
# Load packages:these packages are used for text processing, clustering, and plotting
library(tidyverse)
library(tidytext)
library(proxy)
library(factoextra)
library(ggplot2)

```

Prepare the data and save

```{r}
# Prepare data and create a document ID and clean titles (lowercase)
sample_50 <- sample_50 %>%
  mutate(doc_id = row_number(),
         title = tolower(title))
# Break titles into words, remove stopwords, and compute importance (TF-IDF)
title_tfidf <- sample_50 %>%
  unnest_tokens(word, title) %>%
  anti_join(stop_words) %>%
  filter(str_length(word) > 2) %>%
  filter(!word %in% c("study","analysis","results","review","based","using")) %>%
  count(doc_id, word, sort = TRUE) %>%
  bind_tf_idf(word, doc_id, n)
# Save TF-IDF results
write_csv(title_tfidf, "../output/tfidf_results.csv")
```

Convert into a matrix

```{r}

# Convert data into matrix format for clustering
dtm <- title_tfidf %>%
  tidytext::cast_dtm(doc_id, word, tf_idf)

matrix_data <- as.matrix(dtm)
```

Remove empty rows and measure the similarity coefficient

```{r}

# Remove empty rows (documents with no words)
matrix_data <- matrix_data[rowSums(matrix_data) > 0, ]

# Measure similarity between documents (using cosine distance)
dist_matrix <- proxy::dist(matrix_data, method = "cosine")

```

Group similar documents

```{r}

# Group similar documents into clusters
hc <- hclust(dist_matrix, method = "ward.D2")
```

Plot and save the dendogram.

A dendrogram is a tree diagram used in hierarchical clustering to
visualise the similarity between studies and to guide the selection of
cluster groups.It is a visual guide for cluster not final themes.It
shows structure in the final literature dataset

```{r}

# Plot dendrogram (tree diagram) 
plot(hc, main = "Dendrogram of Study Titles")
# Save plot
png("../output/dendrogram.png", width = 800, height = 600)
plot(hc, main = "Dendrogram of Study Titles")
dev.off()
```

Choose number of clusters

```{r}
# Choose number of clusters (k is defined by you)
k <- 4
clusters <- cutree(hc, k = k)
# Map cluster results back to documents
doc_clusters <- data.frame(
  doc_id = as.numeric(rownames(matrix_data)),
  cluster = clusters
)

```

Merge and save cluster info

```{r}
# Merge cluster info into original data
sample_50 <- sample_50 %>%
  mutate(doc_id = as.integer(doc_id)) %>%
  left_join(doc_clusters, by = "doc_id")
# Save clustered data
write_csv(sample_50, "../output/clustered_data.csv")
```

Find and save the keywords

```{r}
# Find important words in each cluster
cluster_terms <- title_tfidf %>%
  left_join(doc_clusters, by = "doc_id") %>%
  filter(!is.na(cluster)) %>%
  group_by(cluster, word) %>%
  summarise(mean_tf_idf = mean(tf_idf), .groups = "drop") %>%
  arrange(cluster, desc(mean_tf_idf)) %>%
  group_by(cluster) %>%
  slice_head(n = 10)

print(as.data.frame(cluster_terms))
# Save cluster keywords
write_csv(cluster_terms, "../output/cluster_keywords.csv")

```

List the titles by cluster and save the results

```{r}
# List the titles by cluster
result <- sample_50 %>%
  arrange(cluster) %>%
  select(cluster, title)
print(result)
# Save results
write_csv(result, "../output/titles_by_cluster.csv")

```

Check titles inside the clusters

```{r}
# Check titles inside each cluster to help identify the themes
sample_50 %>%
  group_by(cluster) %>%
  summarise(titles = paste(head(title, 5), collapse = " | "))
# We can select the top 3(define as need) from each cluster
cluster_terms %>%
  group_by(cluster) %>%
  summarise(label = paste(head(word, 3), collapse = ", "))

```

Label clusters based on interpretations

```{r}
# Label clusters based on interpretations 
sample_50 <- sample_50 %>%
  mutate(cluster_label = case_when(
    cluster == 1 ~ "Cognitive and Psychological Impacts of Digital Media",
    cluster == 2 ~ "Social Media Use Motivations and Behaviours",
    cluster == 3 ~ "Digital Addiction and Mental Health Outcomes",
    cluster == 4 ~ "Digital Mental Health Tools and Interventions",
    TRUE ~ "Unassigned"
  ))
```

Save labeled data

```{r}
# Save labeled data
write_csv(sample_50, "../output/final_labeled_data.csv")
```

PCA Visualisation

```{r}
# PCA visualisation, profesional for statistical visualisation includes individual as a point
pca <- prcomp(matrix_data, scale. = TRUE)
png("../output/pca_plot.png", width = 800, height = 600)
fviz_pca_ind(
  pca,
  geom = "point",
  habillage = sample_50$cluster_label,
  addEllipses = TRUE,
  title = "Clusters of Studies (PCA)"
)
dev.off()
```

Distribution by publication time

```{r}
# Can also view theme distribution by publication time
p <-ggplot(sample_50, aes(x = year, fill = cluster_label)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Distribution of Studies by Theme Over Time",
    x = "Year",
    y = "Count",
    fill = "Theme"
  ) +
  theme_minimal()
print(p)


```

Save the plot

```{r}
ggsave("../output/time_distribution.png", plot = p, width = 8, height = 6)
```

Facet plot

```{r}
# Facet plot
p2 <- ggplot(sample_50, aes(x = reorder(cluster_label, year), y = year)) +
  geom_point() +
  coord_flip() +
  facet_wrap(~ cluster_label, scales = "free_y") +
  labs(
    title = "Studies Grouped by Theme",
    x = "Title",
    y = "Year"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )+theme(
  axis.text.y = element_text(size = 6)
)

print(p2)

```

Save the plot in the Output folder

```{r}
ggsave(
  "../output/facet_plot.png",
  plot = p2,
  width = 14,     
  height = 10,
  dpi = 300,
  bg = "white"
)
```

# Practice: Prisma table

PRISMA 2020 official site (checklist + flow diagram)

<https://www.prisma-statement.org/>

PRISMA extensions (PRISMA-S for search strategies, PRISMA-P for
protocols)

<https://www.prisma-statement.org/extensions>

```{r }
# load package
library(tidyverse)
```

```{r}
prisma <- data.frame(
  stage = c(
    "Records identified",
    "Duplicates (DOI removed)",
    "Duplicates (clean: title + author)",
    "Manual deduplication",
    "Non-English removed",
    "Before 2016 removed",
    "No abstract available",
    "Title + abstract screened",
    "Records screened",
    "Records excluded",
    "Full-text assessed",
    "Full-text excluded",
    "Studies included"
  ),
  n = c(
    6457,
    694,
    40,
    34,
    nrow(noendata),
    nrow(year_out),
    nrow(no_abstract),
    nrow(with_abstract),
    nrow(combined_data),
    nrow(removed_data),
    nrow(sample_50),
    10,
    40
  )
)
```

Order for flow

```{r}

# Order for flow
prisma$stage <- factor(prisma$stage, levels = rev(prisma$stage))

```

Plot the flow

```{r}
# Flow Plot
p <- ggplot(prisma, aes(x = 1, y = stage)) +
  geom_point(size = 4, color = "steelblue") +
  geom_text(aes(label = paste0(stage, " (n = ", n, ")")),
            hjust = -0.05,
            size = 3.8) +
  geom_segment(aes(x = 1, xend = 1,
                   y = as.numeric(stage),
                   yend = as.numeric(stage) - 1),
               arrow = arrow(length = unit(0.2, "cm")),
               color = "grey50") +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  ) +
  xlim(1, 1.8) +
  ggtitle("PRISMA Flow Diagram (Study Selection Process)")
print(p)
```

Save the Prisma plot

```{r}
ggsave("../output/PRISMA_Flow.png", plot = p, width = 10, height = 8, bg = "white")
```

# Summary for the workshop

Reproducible Systematic Review Project Workflow:

1.  Place database import R
2.  Run scripts in R for cleaning and analysis
3.  Store processed data
4.  Produce reports in reports using R Markdown
5.  Store PRISMA flow diagrams in prisma
