---
title: "Systematic Literature Review Class 1"
author: "Dandian Yang & Lucia Michielin"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Systematic **Literature Review**

Loreum Ipsum @Dandian please enter a short introduction on what will be cover during the first class

## Set up

### Package set up

This part may vary if you are working locally but to get set up in Noteable below you can find all the packages that need to be installed and loaded to run the code.

```{r }
# Install and load the required packages
install.packages(c("synthesisr"))
install.packages("revtools", dependencies = TRUE)
install.packages("bibliometrix", dependencies = TRUE)
install.packages("stringr", " >= 1.5.2") # this is needed because the stringr version in Noteable is too old
install.packages("magrittr", " >= 2.0.4")# this is needed because the magrittr version in Noteable is too old 
#if you are running it locally you may need to run the below as well
#install.packages ("dplyr")
library(revtools)
library(synthesisr)
library(dplyr)
library(bibliometrix)
```

### Import and Merge data

1.  Define the import folder path

```{r }
folder_path <- "../Data/"
```

2.  Import Ris files *N.B.The code below will fail, and it is expected*

```{r}
ref1 <- read_bibliography("../Data/wos1000.ris")
```

2.1 Let's fix the errors

```{r}
lines <- readLines("../Data/wos1000.ris", encoding = "UTF-8")
# check first 20 lines
head(lines, 20)  
# check last 20 lines and we see one empty
tail(lines, 20)  
# remove empty lines
lines <- lines[lines != ""]  
```

2.2 Export the cleaned file

```{r}
writeLines(lines, "../Data/wos1000_clean.ris")
ref1 <- read_bibliography("../Data/wos1000_clean.ris") 
# a preview of column name and type
str(ref1)
```

### Exercise 1

Please practice with the others .ris files and see what you find.

TIPS: There are multiple issues

1.  Weird characters (e.g. smart quotes, accents)
2.  Broken lines
3.  Encoding issues (UTF-8 vs Latin1)
4.  Long abstracts

```{r}
# Space for your code
```

## Practice with SynthesisR Package

@Dandian Can you add some info on SynthesisR package

```{r}
ref <- read_refs("../Data/DiscoverEd.ris")
# generate lists with all ris data, have a preview of the data
files <- list.files(
  "../Data",
  pattern = "\\.ris$",
  full.names = TRUE
)
refs_list <- lapply(files, read_refs)
```

You can import multiple data and use SynthesisR to solve the limitation of RevTools and combine the .Ris data into one.

1.  List all .ris files and import(These are the literature data download from Web of Science and DiscoverEd)

```{r}
risfiles <- list.files(folder_path, pattern = "\\.ris$", full.names = TRUE)
```

2.  Read and combine all files into a list

```{r}
dfris_list  <- lapply(risfiles, function(f) {
  df <- read_refs(f, return_df = TRUE)
  # convert list to character
  df[] <- lapply(df, function(col) {
    if (is.list(col)) {
      sapply(col, function(x) paste(unlist(x), collapse = "; "))
    } else {
      col
    }
  })
  # track source
  df$source_file <- basename(f)
  df
})
```

3.  View the column names under each column to see the number&name of the columns

```{r}
lapply(dfris_list,colnames)
```

4.  View the column type before merge them together

```{r}
lapply(dfris_list, function(df) sapply(df, class))
```

5.  Use bind_rows to combine lists as columns with numbers not the same (18vs19),the missing ones fill with NA

```{r}
combinedris_df <- bind_rows(dfris_list)
colnames(combinedris_df)
```

## Import CSV files

1.  List all CSV files fomr folder path

```{r}
csvfiles <- list.files(folder_path, pattern = "\\.csv$", full.names = TRUE)
```

2.  Read and combine

```{r}
dfcsv_list <- lapply(csvfiles, read.csv)
```

3.  View the column names under each column to see the number&name of the columns

```{r}
lapply(dfcsv_list,colnames)
```

4.  View the column type before merge them together

```{r}
lapply(dfcsv_list, function(df) sapply(df, class))
```

5.  As columns from two datasets are partly match(some overlapping names, different format and different names, we need to standardise name before combine them)

```{r}
clean_names <- function(df) {
  #lowercase all column names
  names(df) <- tolower(names(df))
  #fix the source and source.title issue from list 2
  if (all(c("source.title", "source") %in% names(df))) {
    names(df)[names(df) == "source"] <- "source_file"
    names(df)[names(df) == "source.title"] <- "source"
    
  } else if ("source.title" %in% names(df)) {
    # only source.title exists
    names(df)[names(df) == "source.title"] <- "source"
  }
  #standardise known overlapping columns (as we already lowercase, any contains upercase cannot match)
  names(df)[names(df) %in% c("cites", "cited.by")] <- "citations"
  names(df)[names(df) %in% c("authors")] <- "authors"
  names(df)[names(df) %in% c("title")] <- "title"
  names(df)[names(df) %in% c("year")] <- "year"
  names(df)[names(df) %in% c("doi")] <- "doi"
  names(df)[names(df) %in% c("abstract")] <- "abstract"
  names(df)[names(df) %in% c("page.start","startpage")] <- "start_page"
  names(df)[names(df) %in% c("page.end","endpage")] <- "end_page"
  names(df)[names(df) %in% c("document.type")] <- "source_type"
  return(df)
}
```

6.  Apply to all csv files

```{r}
cleancsv_list <- lapply(dfcsv_list, clean_names)
```

7.  View the column names again

```{r}
lapply(cleancsv_list,colnames)
```

8.  View the column format again

```{r}
lapply(cleancsv_list, function(df) sapply(df, class))
combinedcsv_df <- bind_rows(cleancsv_list)
colnames(combinedcsv_df)
```

## Combine CSV and RIS

View two data frame find commons and differences

```{r}
colnames(combinedcsv_df)
unique(combinedcsv_df$fulltexturl)
colnames(combinedris_df)
```

Keep all the url columns

```{r}
url_cols <- c(
  "articleurl", "citationurl", "citesurl",
  "fulltexturl", "relatedurl", "link"
)
```

List columns has shared structure

```{r}
common_cols <- c(
  "title", "author", "year", "source",
  "publisher", "volume", "issue",
  "start_page", "end_page",
  "doi", "issn", "abstract", "source_file",
  url_cols
)
```

Rename columns to match

```{r}
csv_df <- combinedcsv_df
names(csv_df)[names(csv_df) == "authors"] <- "author"
ris_df <- combinedris_df
# add columns 
add_missing_cols <- function(df, cols) {
  missing <- setdiff(cols, names(df))
  df[missing] <- NA
  df[, cols]
}

csv_df <- add_missing_cols(csv_df, common_cols)
ris_df <- add_missing_cols(ris_df, common_cols)
combined_all <- rbind(csv_df, ris_df)
str(combined_all)
head(combined_all)
```

## Detect duplicates

Finding duplicates from 6457.

Remove duplicates for rows that have a DOI.

```{r}
doi_unique <- combined_all %>%
  filter(!is.na(doi) & doi != "") %>%  # keep only rows with valid DOI
  distinct(doi, .keep_all = TRUE)      # keep the first occurrence of each DOI
```

Keep rows without DOI (NA or empty string)

```{r}
no_doi <- combined_all %>%
  filter(is.na(doi) | doi == "")
```

Combine DOI-unique rows with no-DOI rows

```{r}
combined_unique <- bind_rows(doi_unique, no_doi)
```

Check the result 5770 left

```{r}
nrow(combined_unique)
```

Deduplicate by title for rows without DOI (NA or empty string)

```{r}
no_doi_unique <- combined_unique %>%
  filter(is.na(doi) | doi == "") %>%
  distinct(title, .keep_all = TRUE)
```

Keep DOI rows as is

```{r}
doi_rows <- combined_unique %>%
  filter(!is.na(doi) & doi != "")
```

Combine DOI-unique rows with deduplicated no-DOI rows

```{r}
combined_unique <- bind_rows(doi_rows, no_doi_unique)
```

Check the result 5763 left

```{r}
nrow(combined_unique)
```

Standardise title and author

```{r}
combined_unique <- combined_unique %>%
  mutate(
    doi = ifelse(doi == "", NA, doi),
    title_clean = tolower(trimws(title)),
    author_clean = tolower(trimws(author))
  )
```

Identify groups with the same title + author but multiple DOIs

```{r}
possible_duplicate_doi <- combined_unique %>%
  group_by(title_clean, author_clean) %>%
  filter(n_distinct(doi) > 1) %>%   # only groups with >1 DOI
  mutate(row_number_in_group = row_number()) %>%  # order within group
  ungroup()
```

Keep only first and third rows in each duplicate group

```{r}
keep_rows <- possible_duplicate_doi %>%
  filter(row_number_in_group %in% c(1, 3))
```

Combine with non-duplicate 5761left

```{r}
non_duplicates <- combined_unique %>%
  anti_join(possible_duplicate_doi, by = c("title_clean", "author_clean"))
combined_unique <- bind_rows(non_duplicates, keep_rows) %>%
  arrange(title_clean, author_clean)

```

Deduplicate from title

```{r}
combined_unique$title <- tolower(trimws(combined_unique$title))
combined_final <- combined_unique %>%
  group_by(title_clean) %>%
  slice(1) %>%  # keep only the first row in each group
  ungroup()
removed_rows <- combined_unique %>%
  anti_join(combined_final, by = colnames(combined_unique))
View(removed_rows)  # to review which rows were removed 5730 left
```

## Write the combined data frame to a new CSV file

```{r }
dir.create("../Data/clean")
write.csv(combined_unique, "../Data/clean/metadata.csv", row.names = FALSE)
write_refs(combined_unique, "../Data/clean/metadata.ris", format = "ris")

```

### End of class 1
